\documentclass[11pt]{article}

\usepackage[letterpaper,margin=0.9in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{setspace}
\usepackage{titlesec}
\usepackage{enumitem}
\usepackage[numbers,sort&compress]{natbib}
\usepackage[hidelinks]{hyperref}

\setstretch{1.03}
\setlength{\parindent}{0pt}
\setlength{\parskip}{0.4em}
\setlist[itemize]{leftmargin=1.2em,topsep=2pt,itemsep=1pt,parsep=0pt,partopsep=0pt}
\titlespacing*{\section}{0pt}{0.35em}{0.2em}
\titleformat{\section}{\bfseries\normalsize}{}{0pt}{}
\pagestyle{empty}
\setlength{\bibsep}{1pt}

\begin{document}

\begin{center}
{\large\bfseries Proposal: a zero-inflated negative binomial model for differential expression analysis in single-cell RNA-seq data}\\[-0.2em]
{\normalsize Chenghao Wang \quad | \quad \today}
\end{center}
\vspace{-0.2em}
\hrule
\vspace{0.35em}

\section*{Problem Statement}
With advances in single-cell RNA-sequencing (scRNA-seq) technologies,
the development of statistical approaches for finding differentially
expressed genes become important.

\section*{Background}
A wide variety of tools exist for differential expression (DE)
analysis, ranging from simple non-parametric tests\cite{hao2024dictionary}, through
linear\cite{law_voom_2014}\cite{finak_mast_2015} and generalized linear model\cite{love_moderated_2014}\cite{mccarthy_differential_2012}\cite{miao_desingle_2018} 
based methods, to Bayesian methods
based on mixture models\cite{kharchenko_bayesian_2014}\cite{korthauer_statistical_2016}.

\section*{Data}
We will analyze the dataset by Islam et al.\cite{islam2011characterization}, which is
available at \href{https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE29087}{GSE29087}.
This dataset contains 48 embryonic stems cells and 44 embryonic fibroblasts
in the mouse. The 22,936 features will be filtered, retaining the coding
genes with positive counts in at least 5 cells. We will also simulate data
from zero-inflated negative binomial distributions.
The simulation settings will be designed to reflect the properites
of the dataset by Islam et al.

\section*{Proposed Approach} 
The read counts for each gene will be modeled individually using a zero-inflated negative binomial model.
The normalization will be performed using the same method as in DEsingle\cite{miao_desingle_2018}.
We will use a BFGS algorithm and an generalized EM (GEM) algorithm plus
a Fisher's scoring algorithm in the M-step for estimation,
and we will compare the estimates produced by these two optimizers.
An empirical Bayes method will be used to shrink the gene-wise dispersion estimates.
Likelihood ratio tests and Wald tests will be used to test for DE.

\section*{Evaluation plan}
First, we will apply the proposed method to the Islam et al. dataset
and evaluate the results against their paper.
Next, we will compare the proposed method with MAST\cite{finak_mast_2015} in terms of type I error control by
permuting the condition labels of the cells and assessing the distribution of the p-values.
Lastly, we will compare the proposed method with MAST with respect to ROC curves using
the simulated data.

\section*{Timeline}
\textbf{Week 1}: Derivation of model equations  \\
\textbf{Week 2--3:} Implementation of BFGS \\
\textbf{Week 4--5:} Implementation of GEM + Fisher's scoring \\
\textbf{Week 6:} Implementation of empirical Bayes framework \\
\textbf{Week 7:} Data simulation and method evaluation

{\small
\bibliographystyle{unsrt}
\bibliography{../ref}
}

\end{document}
